---
title: "Market Regime Dynamics and the Role of Investor Sentiment: A Hidden Markov Model Analysis"
author: "Yujia Li"
date: "`r Sys.Date()`" #  
output:
  pdf_document:
    latex_engine: xelatex
spacing: double
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{project Title}
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
- \usepackage{setspace}\doublespacing
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(quantmod)
library(tidyverse)
library(depmixS4)
library(lubridate)
library(dplyr)
library(tibble)
library(tidyr)
library(scales)

suppressPackageStartupMessages(library(quantmod))
```

\newpage

# **1. Introduction**

Financial markets often display regime switching behavior, typically
interpreted as transitions between bull markets and bear markets. These
regimes are not directly observable, but can be inferred from
time-series dynamics. This motivates the use of the Hidden Markov Model
(HMM) to infer latent market states.

Hidden Markov Models (HMMs) provide an intuitive way to uncover the
market patterns we cannot directly see. Following Rabiner’s (1989)
classic formulation, we use tools like the Forward–Backward and Viterbi
algorithms to infer how returns today may reflect underlying bull or
bear conditions. This perspective aligns with Hamilton’s (1989) insight
that financial and macroeconomic series often shift between distinct
regimes, each with its own behavior in terms of average returns and
volatility. To better capture how real-world sentiment shapes these
shifts, we incorporate Filardo’s (1994) time-varying transition
probability framework, which allows optimism, fear, or broader market
signals to influence the likelihood of moving between regimes. All
parameters, like state means, volatilities, and transition dynamics,
which are estimated through the EM-based Baum–Welch procedure (Baum et
al., 1970), ensuring that our identification of bull and bear markets is
grounded in a coherent and data-driven likelihood approach.

Furthermore, the impact of investor sentiment on industry transformation
cannot be ignored. This study considers using the VIX (Volatility Index)
to further infer the potential bull and bear market states in the stock
market by discussing recent fluctuations in AI-related stocks and
indices. This study investigates:

1.  Whether stock market dynamics exhibit statistically distinct bull
    and bear regimes.
2.  Whether overall market sentiment influences the probability of
    transitioning between regimes.
3.  Whether AI-sector-specific sentiment provides additional explanatory
    power in regime transitions.

# **2. Application and Simualtion**

## **2.1 Data Overview:**

This study uses daily financial and sentiment data from **January 1,
2020 to the most recent available date**, including the COVID-19 crash,
the subsequent liquidity-driven recovery, the 2022 tightening cycle, and
the 2023–2024 AI‐led technology rally. Such a sample provides rich
variation for identifying latent market regimes in a Hidden Markov Model
(HMM) framework.

The dataset includes three elements: the S&P 500 (\^GSPC) as the broad
market benchmark for identifying bull–bear regimes; AI-related assets
(NVDA, AMD, SOXX) to capture technology and semiconductor sector
dynamics; and the VIX (\^VIX), which serves as the sentiment indicator
and key driver of transition probabilities in the TVTP model.

## **2.2 Model Specification**

The two-state Gaussian HMM is defined by:

-   **State-dependent return distribution:**
    $r_t \mid S_t=i \sim \mathcal{N}(\mu_i,\sigma_i^2), \quad i \in \{1,2\}$

-   **Transition probability matrix:**
    $p_{ij}​=Pr(S_t​=j∣S_{t−1}​=i), \quad i,j \in \{1,2\}$

-   **Estimation method:** The model parameters $(\mu_i, \sigma_i^2, P)$
    are estimated using the Baum–Welch Expectation–Maximization (EM)
    algorithm, the standard procedure for fitting Hidden Markov Models.
    Because the latent regimes $S_t$​ are unobserved, the likelihood of
    the observed return series $r_t$​ must integrate over all possible
    hidden state paths $(S_1,…,S_T)$. This leads to the complete-data
    likelihood:
    $L(\theta) = \sum_{S_1,\dots,S_T}    P(S_1) \prod_{t=2}^{T}P(S_t|S_{t-1})    \prod_{t=1}^T P(r_t|S_t, \theta)$.
    $P(S_1)$ is the initial regime probability, $P(S_t \mid S_{t-1})$ is
    the regime transition probability, and $P(r_t \mid S_t,\theta)$ is
    the Gaussian emission density under state $S_t$​. The summation
    reflects the fact that the hidden regimes are not observed, so the
    likelihood must consider every possible regime sequence, weighted by
    its probability under the model. Repeating these steps increases the
    likelihood at each iteration until convergence. The summation
    reflects the fact that the hidden regimes are not observed, so the
    likelihood must consider *every possible regime sequence*, weighted
    by its probability under the model. This is allowing the HMM to
    uncover the latent bull and bear regimes that best explain the
    observed return dynamics.

# **3. Data Analysis**

## **3.1 Identifying Bull and Bear Regimes Using a Two-State Hidden Markov Model**

Financial markets often exhibit alternating periods of sustained growth
(“bull markets”) and persistent declines (“bear markets”). These states
are economically meaningful but cannot be directly observed from price
series alone due to noise, short-term volatility, and structural shocks.
HMMs provide a natural probabilistic tool for uncovering such latent
regimes by assuming that returns are generated from state‐dependent
distributions, while the unobserved state follows a first‐order Markov
chain. Following Hamilton’s seminal regime-switching model (1989), U.S.
equity market returns $r_t$ are modeled as:
$$r_t \mid S_t = i \sim \mathcal{N}(\mu_i, \sigma_i^2), \quad i \in \{1,2\}$$where,
$r_t$​ is an unobserved state variable. To estimate the baseline HMM
using **daily log returns of S&P 500 index (\^GSPC).** Computed as:
$r_t = \log(P_t) - \log(P_{t-1}).$

```{r, results='hide', echo=FALSE}
symbols <- c("^GSPC")
getSymbols(symbols, from = "2020-01-01", src = "yahoo")
ret_sp  <- dailyReturn(Cl(`GSPC`), type = "log")
data_sp <- tibble(date = index(ret_sp), r = as.numeric(ret_sp))
set.seed(1234)
mod_sp <- depmix(response = r ~ 1, data = data_sp, nstates = 2, family = gaussian())
fit_sp <- fit(mod_sp)
summary(fit_sp)
```

|             | Mean   | Standard Deviation |
|-------------|--------|--------------------|
| **State 1** | 0.001  | 0.001              |
| **State 2** | -0.002 | 0.025              |

***(Full Summary table see in Appendix A.1)***

```{r, echo=FALSE, out.width="80%"}
post_sp <- posterior(fit_sp, type = "smoothing")
post_df <- as.data.frame(post_sp)
colnames(post_df) <- paste0("S", 1:ncol(post_df))
state_hat <- max.col(post_df)
data_sp <- data_sp |> mutate(state_hat = state_hat,
                             p_state1  = post_df$S1,p_state2  = post_df$S2)
ggplot(data_sp, aes(x = date, y = r, fill = factor(state_hat))) +
  geom_col(alpha = 0.8) + labs(
    title = "S&P 500 Returns with HMM States", 
                               fill = "state_hat") + theme_minimal()
```

Figure 1 (Appendix B.1 includes the full R script): The vertical axis of
this histogram represents **the magnitude of daily returns**, **showing
the daily logarithmic returns** of the S&P 500 index. The colors
indicate the most probable state in the Hidden Markov Model (HMM) for
that day. **A key feature is that the two colors are not randomly
distributed**, but rather form clear, persistent intervals: red
corresponds to periods of low volatility and stability, while blue
concentrates on phases of significantly increased volatility. The blue
histogram is significantly wider, reflecting that the volatility of
state 2 (0.025) is much higher than that of state 1 (0.008), which is
completely consistent with the model parameters and summary statistics
above.

Overall, one state consistently exhibits low mean and high volatility
(bear market characteristics), while the other exhibits high mean and
low volatility (bull market characteristics). Importantly, these labels
are not prior settings, but are naturally derived from parameter
estimates and posterior probabilities. ***Therefore, the results clearly
demonstrate that statistically identifiable and economically significant
bull and bear market states do exist in the US stock market, from a
financial perspective.***

## **3.2 Average and volatility of US stock market returns under different conditions**

Although Section 3.1 reported the estimated state-specific means and
volatility, point estimates alone cannot determine whether these
differences are statistically significant. Therefore, **this section
supplements the estimation results by performing formal hypothesis
testing (using constrained HMMs and likelihood ratio tests).** It
involves assessing whether the two states exhibit significantly
different conditional means and conditional variances.

The fundamental implication of regime-switching models is that returns
in financial markets exhibit distinct distributional features across
latent states. **If the HMM truly identifies meaningful bull and bear
regimes, then their conditional mean returns and conditional
volatilities should differ statistically.**

### **3.2.1 Empirical and Posterior-Weighted Regime Moment Analysis**

In Section 3.1,
$μ1 = 0.001>0>μ2=-0.002, \quad then\quadμ1−μ2\approx0.003$, showing a
sizable shift in return levels across regimes. And
$\sigma_2(=0.008)\approx 3.1 \times \sigma_1(=0.025)$. This confirms
that the **State 2 reflects a higher volatility** which is more than
**triple the daily volatility of State 1**.

To evaluate whether the two market states identified by the Hidden
Markov Model truly correspond to different market conditions, this
section will use hard classification and posterior-weighted methods to
verify whether bull and bear market states have significant economic
implications.

This method will not rely on the likelihood function, but instead group
observations according to state and calculate the sample mean and
standard deviation within each group.

```{r}
# Sample Statistics After Classifying Each Day into a Regime
reg_summary <- data_sp %>% group_by(state_hat) %>%
  summarise(mean_r = mean(r), sd_r = sd(r), n = n())
reg_summary <- reg_summary %>%
  mutate(mean_annual = mean_r*252,sd_annual = sd_r*sqrt(252),
         pct_time = n/sum(n)*100)
reg_summary
```

To validate the model inference from a data perspective, a weighted mean
was further calculated using posterior probabilities:

$$\hat{\mu}_i = \frac{\sum_t p_{ti} \, r_t}{\sum_t p_{ti}}, \quad \hat{\sigma}_i^2 = \frac{\sum_t p_{ti} (r_t - \hat{\mu}_i)^2}{\sum_t p_{ti}}$$

```{r}
# Posterior-Weighted Empirical Estimates
data_sp %>% summarise(mean_state1 = weighted.mean(r, p_state1),
                      mean_state2 = weighted.mean(r, p_state2),
            sd_state1 = sqrt(weighted.mean((r)^2, p_state1) - mean_state1^2),
            sd_state2   = sqrt(weighted.mean((r)^2, p_state2) - mean_state2^2))
```

The parameter estimation of Hidden Markov Models (HMMs) provides
likelihood-based state characteristics, describing how the model
theoretically distinguishes different return distributions through mean
and volatility. To validate these characteristics from a data
perspective, further employ **hard classification**, assigning each day
to its most probable state and calculating the sample mean and standard
deviation **to examine whether the grouped observations show significant
differences in the actual data**. Simultaneously, **posterior
weighted**, by preserving the state probability at each time point,
avoid forced 0/1 classification and **provide a more flexible state
characterization method consistent with the model structure.** These
three methods together constitute a validation framework for regime
characteristics from the three levels of model, data, and probability.

**It is evident that the mean and variance obtained by the three methods
are very similar,** indicating robust model estimation and clear state
identification. State 1 (bull market) consistently exhibits higher
average returns and lower volatility, while State 2 (bear market)
exhibits lower returns and higher volatility. This suggests that the
inferred two phases represent economically meaningful bull and bear
market states, respectively. ***This high degree of consistency
validates the model's effectiveness in distinguishing between bull and
bear market states, and also indicates that there are indeed significant
differences in the mean and volatility between the two phases.***

### **3.2.2 Likelihood Ratio Tests（LRT） for Regime Separation**

To test whether the mean/variance are significantly different. Consider
a restricted model: force $μ₁ = μ₂$ (same mean), and compare it with the
original model:
$$LR = 2(\ell_{\text{unconstrained}} - \ell_{\text{restricted}}),$$if
the LR value is significant, it indicates that the states are indeed
different.(Full code see in Appendix A.3)

```{r, include=FALSE}
# constrained model with common mean
mod_equal_mu <- depmix(response = r ~ 1, data = data_sp, 
                       nstates = 2, family = gaussian())
params <- getpars(fit_sp)
params[c(1, 3)] <- mean(params[c(1, 3)])
mod_equal_mu <- setpars(mod_equal_mu, params)
fit_equal_mu <- fit(mod_equal_mu)

# LR test
LR <- 2 * (logLik(fit_sp) - logLik(fit_equal_mu))
p_value <- 1 - pchisq(LR, df = 1) 
p_value
```

**The likelihood ratio test strongly rejects the null hypothesis of
equal means between the two regimes (p = 0.009 \< 0.01),** **and
conclude that the average returns differ significantly.** In terms of
volatility, the bear regime exhibits a standard deviation roughly three
times larger than that of the bull regime. This large gap is clearly
visible in the descriptive statistics and histograms, so we do not need
run a separate formal LRT for equal variances again.

### **3.2.3 Visualization of Regime-Dependent Distributions**

The histogram in the chart uses different colors to show the
distribution of daily returns of the S&P 500 in two implied states.

```{r, echo=FALSE, out.width="80%"}
ggplot(data_sp, aes(x = r, fill = factor(state_hat))) + 
  geom_histogram(alpha = 0.4, bins = 60, position = "identity") + 
  labs(title = "Regime-Dependent Distributions of S&P 500 Returns", 
       fill = "State") + theme_minimal()
```

-   **Red: State 1 (Bull Market)**: shows a more concentrated
    distribution and lower volatility, mainly clustered in a narrow area
    with returns close to zero, indicating lower daily return
    volatility. This aligns with the model's identification of a "stable
    upward trend."

-   **Blue: State 2 (Bear Market)**: shows a more dispersed distribution
    and more extreme values ​​at both ends of the return spectrum
    (especially on the negative return side), indicating that the market
    in this state experiences higher volatility and larger declines.

Figure 2 (Appendix B.2 includes the full R script): ***The histogram
visually illustrates the significant structural differences between bull
and bear markets in terms of return direction and risk level.*** The
estimated HMM parameters, posterior-weighted statistics, and likelihood
ratio tests all indicate that the two inferred regimes are statistically
distinct in both mean and volatility. These differences are economically
intuitive and align with the definitions of bear and bull markets in the
literature. ***Therefore, the U.S. equity market exhibits clear
regime-dependent distributional characteristics, validating the HMM’s
ability to separate meaningful latent market states.***

## **3.3 Does Investor Sentiment Drive Transition Probabilities?**

Consistent with Sections 3.1 & 3.2, $S_t \in \{1,2\}$ denote the latent
regime at time t, where state 1 is bear regime and state 2 is bull
regime. The conditional return distribution remains Gaussian as in the
baseline model:
$r_t \mid S_t = i \sim \mathcal{N}(\mu_i, \sigma_i^2), \quad i \in \{1,2\}$.

### **3.3.1 TVTP-HMM with VIX**

**The key innovation of TVTP-HMM lies in the fact that the transition
probabilities are no longer constant.** **It models the transition
probabilities as a function of the standardized sentiment index**
$Z_{t-1}$**,** based on Filardo's (1994) covariate dependency transition
specification and the formula summarized by Zucchini et al. (2016):
$\text{logit}\big(p_{11}(t)\big) = \alpha_{10} + \alpha_{11} Z_{t-1},\quad \text{logit}\big(p_{22}(t)\big) = \alpha_{20} + \alpha_{21} Z_{t-1}$

**Economic Intuition** **for the coefficients are:**

-   $\alpha_{11} < 0$ : higher VIX (more fear) should **reduce the
    persistence of bull markets**

-   $\alpha_{21} < 0$ : higher VIX should **increase the probability of
    exiting bull markets into bear markets**

```{r, warning=FALSE, results='hide', echo=FALSE}
# Download data
getSymbols("^VIX", src = "yahoo", from = "2020-01-01")
vix <- Cl(`VIX`)
vix_df <- tibble(date = index(vix), vix  = as.numeric(vix)) %>% na.omit
# Merge with S&P 500 returns data (data_sp from the previous)
data_tv <- data_sp %>% inner_join(vix_df, by = "date") %>%
  mutate(vix_std  = as.numeric(scale(vix)),
         vix_lag1 = dplyr::lag(vix_std, 1)) %>% drop_na(r, vix_lag1)
# TVTP-HMM: transition depends on lagged VIX
set.seed(1234)
mod_tvtp <- depmix(response = r ~ 1, data = data_tv, 
                   nstates = 2, family = gaussian(), transition = ~ vix_lag1)
fit_tvtp <- fit(mod_tvtp)
summary(fit_tvtp)
```

| Transition (i → j)    | Intercept | Coefficient on lagged VIX (vix_lag1) |
|-----------------------|-----------|--------------------------------------|
| **State 1 → State 2** | −2.618516 | **3.183426**                         |
| **State 2 → State 2** | −1.668077 | **1.481982**                         |

***(Full Summary table see in Appendix A.3)***

In the transition model corresponding to a low-volatility bull market
(State 1), the VIX coefficient is **positive and relatively large
(3.18)**. This means that increased market panic reduces the
sustainability of the bull market, making it easier for the market to
enter a bear market. The VIX coefficient of bull market (State 2) is
positive at 1.48, which indicates the stronger the panic, the harder it
is for the market to escape the bear market. Higher VIX levels enhance
the persistence of the bear market while reducing the probability of the
market re-entering a bull market. **The estimated TVTP-HMM provides
strong evidence that investor sentiment, proxied by lagged VIX, plays a
significant role in driving regime transitions.**

To verify the relationship between the probability of a bull market and
the VIX index, we visualize the relationship. To set the **blue line as
the probability of a bull market**, the **red line as the VIX index**,
and the **x-axis as dates**.

```{r, warning=FALSE, out.width="80%", echo=FALSE}
post_tvtp <- posterior(fit_tvtp, type = "smoothing") %>% as.data.frame()
colnames(post_tvtp) <- c("p_state1", "p_state2")
data_tv <- data_tv %>% mutate(p_bull_tvtp = post_tvtp$p_state2)
ggplot(data_tv, aes(x = date)) + 
  geom_line(aes(y = p_bull_tvtp), color = "blue", size = 0.3) +
  geom_line(aes(y = rescale(vix, to = c(0, 1))),color = "red", 
            linetype = "dashed", size = 0.5) + labs(
              title = "Bull Regime Probability vs. VIX",
              x = NULL, y = "Probability / Rescaled VIX") + theme_minimal()
```

Figure 3 (Appendix B.3 includes the full R script): The visual evidence
shows a sharp and systematic inverse comovement between the bull-regime
probability and the VIX index. First, **spikes in VIX** (e.g., during
the COVID-19 crash, the 2022–2023 monetary tightening cycle, or other
episodes of macro uncertainty) tend to coincide with sharp **declines in
the bull-regime probability** and more frequent switches into the bear
regime. This consistents with the idea that fear accelerates regime
changes. Second, during calmer periods with low VIX, the bull regime
becomes more persistent, and the state classification from the TVTP
model is more stable than under the constant-transition specification.

Overall, the VIX not only reflects market volatility but is also a
significant driver of market state transitions. Its changes are highly
correlated with investor sentiment, and it accelerates the market's
shift towards a bear market when uncertainty rises. This pattern aligns
with the separation of mean and variance in the benchmark HMM, further
demonstrating the robust economic and statistical significance of
bull-bear market segmentation.

### **3.3.2 Case Study: AI-Linked Semiconductor Volatility**

The recent surge in artificial intelligence illustrates sentiment-driven
market dynamics. Since the end of 2022, semiconductor stocks related to
generative AI, particularly sector ETFs such as **Nvidia (NVDA), AMD,
and SOXX**, have experienced exceptionally strong gains, followed by
significant volatility:

-   The Philadelphia Semiconductor Index (SOX), a key benchmark for the
    sector, **more than tripled** between 2020 and 2024, largely driven
    by AI-related demand (Moomoo, 2024). Nvidia’s market value briefly
    exceeded **USD 4 trillion in July 2025**, after its stock price
    increased roughly tenfold since early 2023, reflecting extreme
    investor optimism about AI data-center chips.

**Multiple reports indicate rising market concerns about valuation risks
and industry adjustments**: for example, Vanguard Group warned that the
AI ​​rally may be subject to a correction due to overly high expectations,
and research also shows that volatility and bubble signs in AI-related
stocks are intensifying (Hughes, 2024). **Short-term volatility has
become more frequent:** the stock prices of Nvidia and AMD fluctuate
wildly with earnings reports and news from competitors, the stock price
drops by double digits after news reports about major customers testing
alternative AI chips (Rose, 2025).

***This path of "excessive optimism followed by sharp fluctuations or
even a correction" is a typical manifestation of the state switching
phenomenon that fear indicators such as VIX and TVTP-HMM focus on.***

Based on the existing data_tv (includes S&P 500 returns, VIX, and
p_bull_tvtp), we will construct an additional simplified "AI
Semiconductor Basket Returns" code, based on the equal-weighted average
returns of NVDA, AMD, and SOXX, and then plot it together with the bull
market probability and VIX.

```{r, include=FALSE}
# Download & preprocessing data
symbols_ai <- c("NVDA", "AMD", "SOXX")
getSymbols(symbols_ai, src = "yahoo", from = "2020-01-01")

prices_ai <- bind_rows(tibble(date = index(Ad(NVDA)), symbol = "NVDA",
                              price = as.numeric(Ad(NVDA))),
  tibble(date = index(Ad(AMD)), symbol = "AMD", price = as.numeric(Ad(AMD))),
  tibble(date  = index(Ad(SOXX)), symbol = "SOXX", price  = as.numeric(Ad(SOXX))))

returns_ai <- prices_ai %>% arrange(symbol, date) %>%    
  group_by(symbol) %>% mutate(log_price = log(price),
                              ret = log_price - dplyr::lag(log_price)) %>%
  ungroup() %>% drop_na(ret)

ai_basket <- returns_ai %>% group_by(date) %>%
  summarise(ret_ai_eqw = mean(ret),.groups = "drop")

data_tv_ai <- data_tv %>% inner_join(ai_basket, by = "date") %>%
  mutate(ret_ai_std = as.numeric(scale(ret_ai_eqw)))
```

```{r, warning=FALSE, out.width="80%", echo=FALSE}
ggplot(data_tv_ai, aes(x = date)) +
  geom_line(aes(y = p_bull_tvtp,   color = "Bull probability")) +
  geom_line(aes(y = rescale(vix),  color = "VIX")) +
  geom_line(aes(y = rescale(ret_ai_std), color = "AI returns (scaled)")) +
  scale_color_manual(values = c("Bull probability" = "skyblue", 
                                "VIX" = "red", "AI returns (scaled)" = "orange")) +
  labs(title = "Bull Regime Probability, VIX, and AI-Sector Returns",
       x = NULL, y = "Scaled values (0–1)", color = "") + theme_minimal()
```

Figure 4 (Appendix B.4 includes the full R script): **The visualization
reveals a clear transmission mechanism linking AI-sector volatility,
market sentiment, and regime switching.** Periods of heightened AI-stock
turbulence coincide with spikes in the VIX and sharp collapses in the
bull-regime probability, consistent with the TVTP-HMM estimates that
investor **fear accelerates transitions into the bear regime**. This
pattern is particularly pronounced during the 2022 tightening cycle and
the recent AI-driven volatility episodes.

The TVTP-HMM results complement the earlier findings in Sections 3.1 and
3.2, which established that U.S. equity returns can be meaningfully
decomposed into bull and bear regimes with distinct mean–variance
profiles. Section 3.3 confirmed that these differences are statistically
robust and not an artifact of overfitting a single-regime model.
Overall, the visual evidence is in line with the regression-based
results, and supports the hypothesis that **investor sentiment drives
transition probabilities rather than only affecting within-regime
volatility**.

# **4. Conclusion**

This study aims to examine whether statistically identifiable bull and
bear market mechanisms exist in the US stock market, and whether these
mechanisms exhibit significant differences in returns and volatility. It
further explores whether investor sentiment drives state transitions.

All models exhibit highly consistent patterns. One class of states is
characterized by high returns and low volatility, while the other is
characterized by low returns and high volatility. These patterns
originate from the data itself, rather than any assumptions, which
assures us that the "bull market" and "bear market" labels effectively
reflect different underlying market conditions. The timeline of
posterior probabilities also closely aligns with several well-known
events: the sharp volatility of 2020, the uncertainty of the tightening
cycle in 2022-2023, and the renewed optimism at the end of 2023 driven
by the growing momentum of AI-related industries.

By incorporating investor sentiment, we gain deeper insights. When we
allow state transition probabilities to respond to the lagged VIX index,
periods of higher perceived risk are more likely to see bear market
transitions (or continuations). This result aligns with the intuition
that emotion-driven fear and uncertainty push markets towards more
volatile states.

Overall, Hidden Markov Models provide a powerful framework for
understanding market evolution over time. The results demonstrate that
identifying market mechanisms and the emotional forces behind them helps
in better understanding the shifts in risk, macroeconomic shocks, and
market sentiment.

\newpage

# References

-   **Baum, L. E., Petrie, T., Soules, G., & Weiss, N. (1970). A
    maximization technique occurring in the statistical analysis of
    probabilistic functions of Markov chains. *The Annals of
    Mathematical Statistics*, *41*(1), 164–171.
    <https://doi.org/10.1214/aoms/1177697196>**

-   **Filardo, A. J. (1994). Business-Cycle Phases and their
    transitional dynamics. *ideas.repec.org*.
    <https://ideas.repec.org/a/bes/jnlbes/v12y1994i3p299-308.html>**

-   **Hamilton, J. D. (1989). A new approach to the economic analysis of
    nonstationary time series and the business cycle. *Econometrica*,
    *57*(2), 357. <https://doi.org/10.2307/1912559>**

-   **Hughes, J. (2024, December 4). Wall Street’s AI-powered rally
    risks ‘correction’, Vanguard warns. *Financial Times*.
    <https://www.ft.com/content/8b5c4611-0957-4bca-a6e9-ad608c865a64>**

-   **Moomoo. (2024, July 12). *How to choose among the top 10
    semiconductor ETFs amid the AI
    wave?*<https://www.moomoo.com/my/learn/detail-semiconductor-etfs-surge-comparing-the-top-10-117239-240770101>**

-   **Rabiner, L. (1989). A tutorial on hidden Markov models and
    selected applications in speech recognition. *Proceedings of the
    IEEE*, *77*(2), 257–286. <https://doi.org/10.1109/5.18626>**

-   **Rose, E. (2025, November 25). *AI Stock market uncertainty after
    Nvidia earnings: Is the bubble at risk?* AI Stock Market Uncertainty
    After Nvidia Earnings: Is the Bubble at Risk?
    <https://www.markets.com/analysis/ai-stock-market-uncertainty-after-nvidia-earnings-2665-en>**

\newpage

# **Appendix**

## Appendix A: Supplementary Tables

-   ***Appendix A.1***

```{r}
symbols <- c("^GSPC")
getSymbols(symbols, from = "2020-01-01", src = "yahoo")
ret_sp  <- dailyReturn(Cl(`GSPC`), type = "log")
data_sp <- tibble(date = index(ret_sp), r = as.numeric(ret_sp))
set.seed(1234)
mod_sp <- depmix(response = r ~ 1, data = data_sp, nstates = 2, family = gaussian())
fit_sp <- fit(mod_sp)
summary(fit_sp)
```

-   ***Appendix A.2***

```{r, warning=FALSE}
# Download data
getSymbols("^VIX", src = "yahoo", from = "2020-01-01")
vix <- Cl(`VIX`)
vix_df <- tibble(date = index(vix), vix  = as.numeric(vix)) %>% na.omit
# Merge with S&P 500 returns data (data_sp from the previous)
data_tv <- data_sp %>% inner_join(vix_df, by = "date") %>%
  mutate(vix_std  = as.numeric(scale(vix)),
         vix_lag1 = dplyr::lag(vix_std, 1)) %>% drop_na(r, vix_lag1)
# TVTP-HMM: transition depends on lagged VIX
set.seed(1234)
mod_tvtp <- depmix(response = r ~ 1, data = data_tv, 
                   nstates = 2, family = gaussian(), transition = ~ vix_lag1)
fit_tvtp <- fit(mod_tvtp)
summary(fit_tvtp)
```

-   ***Appendix A.3***

```{r}
# constrained model with common mean
mod_equal_mu <- depmix(response = r ~ 1, data = data_sp, 
                       nstates = 2, family = gaussian())
params <- getpars(fit_sp)
params[c(1, 3)] <- mean(params[c(1, 3)])
mod_equal_mu <- setpars(mod_equal_mu, params)
fit_equal_mu <- fit(mod_equal_mu)

# LR test
LR <- 2 * (logLik(fit_sp) - logLik(fit_equal_mu))
p_value <- 1 - pchisq(LR, df = 1) 
p_value
```

## Appendix B: R Code

-   ***Appendix B.1***

```{r}
# Figure 1: S&P 500 Returns with HMM States
post_sp <- posterior(fit_sp, type = "smoothing")
post_df <- as.data.frame(post_sp)
colnames(post_df) <- paste0("S", 1:ncol(post_df))
state_hat <- max.col(post_df)
data_sp <- data_sp |> mutate(state_hat = state_hat,
                             p_state1  = post_df$S1,p_state2  = post_df$S2)
ggplot(data_sp, aes(x = date, y = r, fill = factor(state_hat))) +
  geom_col(alpha = 0.8) + labs(title = "S&P 500 Returns with HMM States", 
                               fill = "state_hat") + theme_minimal()
```

-   ***Appendix B.2***

```{r}
# Figure 2: Regime-Dependent Distributions of S&P 500 Returns
ggplot(data_sp, aes(x = r, fill = factor(state_hat))) + 
  geom_histogram(alpha = 0.4, bins = 60, position = "identity") + 
  labs(title = "Regime-Dependent Distributions of S&P 500 Returns", 
       fill = "State") + theme_minimal()
```

-   ***Appendix B.3***

```{r}
# Figure 3: Bull Regime Probability vs. VIX
post_tvtp <- posterior(fit_tvtp, type = "smoothing") %>% as.data.frame()
colnames(post_tvtp) <- c("p_state1", "p_state2")
data_tv <- data_tv %>% mutate(p_bull_tvtp = post_tvtp$p_state2)
ggplot(data_tv, aes(x = date)) + 
  geom_line(aes(y = p_bull_tvtp), color = "blue", size = 0.3) +
  geom_line(aes(y = rescale(vix, to = c(0, 1))),color = "red", 
            linetype = "dashed", size = 0.5) + labs(
              title = "Bull Regime Probability vs. VIX",
              x = NULL, y = "Probability / Rescaled VIX") + theme_minimal()
```

-   ***Appendix B.4***

```{r, eval = FALSE}
# Download & preprocessing data
symbols_ai <- c("NVDA", "AMD", "SOXX")
getSymbols(symbols_ai, src = "yahoo", from = "2020-01-01")

prices_ai <- bind_rows(tibble(date = index(Ad(NVDA)), symbol = "NVDA",
                              price = as.numeric(Ad(NVDA))),
  tibble(date = index(Ad(AMD)), symbol = "AMD", price = as.numeric(Ad(AMD))),
  tibble(date  = index(Ad(SOXX)), symbol = "SOXX", price  = as.numeric(Ad(SOXX))))

returns_ai <- prices_ai %>% arrange(symbol, date) %>%    
  group_by(symbol) %>% mutate(log_price = log(price),
                              ret = log_price - dplyr::lag(log_price)) %>%
  ungroup() %>% drop_na(ret)

ai_basket <- returns_ai %>% group_by(date) %>%
  summarise(ret_ai_eqw = mean(ret),.groups = "drop")

data_tv_ai <- data_tv %>% inner_join(ai_basket, by = "date") %>%
  mutate(ret_ai_std = as.numeric(scale(ret_ai_eqw)))
```

```{r}
# Figure 4: Bull Regime Probability, VIX, and AI-Sector Returns
ggplot(data_tv_ai, aes(x = date)) +
  geom_line(aes(y = p_bull_tvtp,   color = "Bull probability")) +
  geom_line(aes(y = rescale(vix),  color = "VIX")) +
  geom_line(aes(y = rescale(ret_ai_std), color = "AI returns (scaled)")) +
  scale_color_manual(values = c("Bull probability" = "skyblue", 
                                "VIX" = "red", "AI returns (scaled)" = "orange")) +
  labs(title = "Bull Regime Probability, VIX, and AI-Sector Returns",
       x = NULL, y = "Scaled values (0–1)", color = "") + theme_minimal()
```
